The Way Forward in Parallelizing Dynamic Languages

### 2014-05-01 talk/icooolps2014/position-paper.tex
say what we are working on

The GIL also allows for easy integration with external C libraries that
do not need to be thread-safe. For the duration of the calls, we
simply do not release the GIL. External libraries that are explicitly
thread-safe can voluntarily release the GIL themselves in order to
still provide some parallelism. This is done for example for
potentially long I/O operations. Consequently, I/O-bound,
multi-threaded applications can actually parallelise to some
degree. Again, a potential solution should be able to integrate with
external libraries with similar ease. We will however focus our
argumentation more on running code in the interpreted language in
parallel, not the external C calls.

@@

TM can be implemented in software (STM) or in hardware (HTM. There are
also some hybrid approaches that combine the two. We count these
hybrid approaches as STM, since they usually provide the same
capabilities as software-only approaches but with different
performance characteristics. We will now first look at HTM that
recently gained a lot of popularity by its introduction in common
desktop CPUs from Intel (Haswell generation).

@@

Following the above argumentation for each approach we assembled a
general overview in Table \ref{tab:comparison}. The general picture is
everything else than clear. It looks like HTM may be a good solution
to replace the GIL in the future. Current implementations are however
far too limiting and do not provide good scaling.

Just allowing for parallel execution only means that dynamic languages
catch up to all other languages that already provide real
parallelism. This is why we think that only the STM approach is a
viable solution in the long-term. It provides the application with a
simple memory model (sequential consistency) and a composable way to
synchronise memory accesses using atomic blocks.

STM has a big performance problem. We believe that further work
to reduce the overhead by closely working together with the
hardware should be the focus of research. Hybrid approaches that
combine STM and HTM for performance may be able to overcome this
obstacle.

>>>>

The GIL also allows for easy integration with external C libraries that
may not be thread-safe. For the duration of the calls, we
simply do not release the GIL. External libraries that are explicitly
thread-safe can voluntarily release the GIL themselves in order to
still provide some parallelism. This is done for example for
potentially long I/O operations. Consequently, I/O-bound,
multi-threaded applications can actually parallelise to some
degree. Again, a potential solution should be able to integrate with
external libraries with similar ease. We will however focus our
argumentation more on running code in the interpreted language in
parallel, not the external C calls.

@@

TM can be implemented in software (STM) or in hardware (HTM). There
are also hybrid approaches, which combine the two. We count these
hybrid approaches as STM, since they usually provide the same
capabilities as software-only approaches but with different
performance characteristics. We will now first look at HTM, which
recently gained a lot of popularity by its introduction in common
desktop CPUs from Intel (Haswell generation).

@@

Following the above argumentation for each approach we assembled a
general overview in Table \ref{tab:comparison}. The general picture is
everything else than clear. It looks like HTM may be a good solution
to replace the GIL in the near future. Current implementations are
however far too limiting and do not provide good scaling.

Allowing for parallel execution just means that dynamic languages
catch up to all other languages that already provide real
parallelism. This is why we think that only the STM approach is a
viable solution in the long-term. It provides the application with a
simple memory model (sequential consistency) and a composable way to
synchronise memory accesses using atomic blocks.

Unfortunately, STM has a big performance problem. One way to approach
this problem is to make STM systems that use the available hardware
better. We are currently working on a STM system that makes use of
several hardware features like virtual memory and memory segmentation.
We further tailor the system to the discussed use case which gives us
an advantage over other STM systems that are more general.  With this
approach, initial results suggest that we can keep the overhead of STM
already below 50\%. A hybrid TM system, which also uses HTM to
accelerate certain tasks, looks like a very promising direction of
research too. In general we believe that further work to reduce the
overhead of STM is very worthwhile.
